Formalizing the Edmonds-Karp Algorithm

### 2016-03-06 paper/root.tex
Not measuring initial tabulation on SML benchmark any more
3rd iteration over paper
Added introduction to "formalizing Fofu" section
Added "implementing augmentation" subsection
Updated benchmarks. Diagrams now next to each other. (I'm not yet completely satisfied with the result)

\author{S.~Reza Sefidgar, Peter Lammich}

\institute{Technische Universit\"at M\"unchen, \email{\{sefidgar,lammich\}@in.tum.de}}

@@

\section{Introduction}
Computing the maximum flow of a network is an important problem in graph theory.
Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, and various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem, an important instance being the Edmonds-Karp algorithm~\cite{EK72},
which was one of the first algorithms to solve the maximum flow problem in polynomial time for the general case of real valued capacities.

@@

While there exists another formalization of the Ford-Fulkerson method~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation.

@@

To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction; and 3) every vertex of the flow network must be on a path from $s$ to $t$. Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.

@@

The Ford-Fulkerson method is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose the Ford-Fulkerson method defines the residual graph, which has edges both in parallel and reverse to the edges of the network. 
Each edge is labeled by the amount of flow that can be effectively moved along this edge, by either increasing or decreasing the flow on an edge in the network.
Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

@@

The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph, and there are several possible implementations with different execution times. The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}. When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real valued capacities. Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges.
A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.

\section{Formalizing the Ford-Fulkerson method}\label{sec:abs-formalization}
%  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
%    Reader should think: Yeah, they did cool stuff!
%
%  [???* Thematize definition of flows: On Graphs vs. on Networks. You find both in the literature. 
%      We decides for [...]. This is better suited, as is gives nice and elegant proofs, even formally, bla bla bla]
%    
%  Proof of MinCut-MaxFlow: Follows the textbook proof.
%    Uses Isar to even look like a textbook proof, being comprehensible even without running
%      Isabelle/HOL.
%      
%    For example, \isai{augment_flow_presv_cap}.
%      Present formal proof text. Perhaps oppose it to textbook proof?.
%    
%  Abstract Algo looks like pseudocode presented in texbooks.
%    Oppose our algo to textbook pseudocode.
%

@@

% (DONE: not sure about repeating citation of Cormen: I would definitely repeat the citation!) 
Informal proofs focus on the relevant thoughts by leaving out technical details and obvious steps. 
In contrast, a formal proof has to precisely specify each step as the application of some inference rules. 
Although modern proof assistants provide high-level tactics to summarize some of these steps, formal proofs tend to be significantly more 
verbose than informal proofs. Moreover, formal proofs are conducted in the tactic language of the proof assistant, which is often some dialect of ML. 
Thus, many formal proofs are essentially programs that instruct the proof assistant how to conduct the proof. They tend to be inaccessible without a deep knowledge of the used proof assistant, in many cases requiring to replay the proof in the proof assistant in order to understand the idea behind it.

@@

In textbooks, it is common to present algorithms in pseudocode, which captures the essential ideas, but leaves open implementation details. As a formal equivalent to pseudocode, we use the monadic programming language provided by the Isabelle Refinement Framework~\cite{LaTu12,La12}:
\Snippet{ford_fulkerson_algo}

% DONE: Introduce obtain-syntax, and unfold find-augmenting-path spec! Or even while-exists loop as combinator?
The code looks quite similar to pseudocode that one would expect in a textbook, but actually is a rigorous formal specification of the algorithm, using nondeterminism to leave open the implementation details (\cf Section~\ref{sec:refinement}). Note that we had to use the available combinators of the Isabelle Refinement Framework, which made the code slightly more verbose than we would have liked. We leave it to future work to define a set of combinators and appropriate syntax that allow for more concise presentation of pseudocode.

@@

In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12,La12}. 
It features a refinement calculus for programs phrased in a nondeterminism monad. 
The monad's type is a set of possible results plus an additional value that indicates a failure:
\begin{lstlisting}
  datatype \<alpha> nres = res \<alpha> set | fail
\end{lstlisting}
The operation \isai{return x} of the monad describes the single result \isai{x}, and the 
operation \isai{bind m f} nondeterministically picks a result from $m$ and executes $f$ on it. 
If either \isai{m = fail}, or \isai{f} may fail for a result in $m$, the bind operation fails.

We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
Intuitively, $m\le m'$ means that $m$ is a refinement of $m'$, \ie all possible results of $m$ are 
also possible results of $m'$. 
Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define standard loop constructs (while, foreach), 
a syntax for postcondition specifications, and use a Haskell-like do-notation:
\begin{lstlisting}
  spec P == spec x. P x == res {x. P x}
  do {x <- m; f x} == bind m f
  do {m; m'} == bind m (%_. m')
\end{lstlisting}

Correctness of a program $m$ with precondition $P$ and postcondition $Q$ is expressed as \isai{P ==> m <= spec r. Q r} (or, eta-contracted, just \isai{spec Q}), which
means that, if $P$ holds, $m$ does not fail and all possible results of $m$ satisfy $Q$. Note that we provide different recursion constructs
for partial and total correctness: A nonterminating total correct recursion yields \isai{fail}, which satisfies no specification, even if joined 
with results from other possible runs. On the other hand, a nonterminating partial correct recursion yields \isai|res {}|, which refines any specification 
and disappears when joined with other results.

The Isabelle Refinement Framework also supports data refinement, changing the representation of results according to a \emph{refinement relation}, 
which relates concrete with abstract results: Given a relation $R$, \isai{\<Down>R m} is the largest set of concrete results that are related to an 
abstract result in $m$ by $R$. If \isai{m=fail}, then also \isai{\<Down>R m = fail}.

In a typical program development, one first comes up with an initial version $m_0$ of
the algorithm and its specification $P,Q$, and shows \isai{P ==> m_0 <= spec Q}. Then, one iteratively provides refined versions $m_i$ of the algorithm,
proving \isai{$m_i$ <= \<Down>$R_i$ $m_{i-1}$}. Using transitivity and composability of data refinement, one 
gets \isai{P ==> $m_i$ <= \<Down>$R_i\ldots R_1$ spec Q}, showing the correctness of the refined algorithm.
If no data refinement is performed, $R_i$ is set to the identity relation, in which case \isai{\<Down>$R_i$} becomes the identity function.

@@

\section{The Edmonds-Karp Algorithm}\label{sec:edka}
  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm requires us to show that finding a shortest augmenting path is a refinement of finding any augmenting path, which is straightforward using basic graph theory.

@@

  Formalizing the above intuitive argument was more tricky than it seemed on first glance:
  While it is easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
  to a graph transformation which may add reverse edges and removes at least one original edge requires some generalization of the statement.
  Note that a straightforward induction on the length of the augmenting path or on the number of flipped edges fails, as, after flipping the first 
  edge, the path is no longer augmenting.

@@

  Having defined the measure function, and shown that it decreases on augmentation, it is straightforward to refine the partial correct while loop to a total correct one. Moreover, to make explicit the bound on the number of loop iterations, we instrument the loop with a counter, and assert the upper bound on the counter after the loop.

\section{Refinement to Executable Code}\label{sec:executable}
  In the previous section, we have presented our abstract formalization of the Edmonds-Karp algorithm,
  leaving open how to obtain a shortest augmenting path, and how to implement the algorithm. 
  In this section, we outline the further refinement steps that were necessary to obtain an efficient implementation.

@@

@@

  We implement capacities as (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume real, rational, and integer values.}. Note that an implementation as fixed precision numbers would also be possible,
  but requires additional checks on the network to ensure that no overflows can happen. 
  % TODO: Can we trick the Java reference implementation on integer overflow?

@@

  The input network of the algorithm is represented as a function from network edges to capacities, which is tabulated into an array
  to obtain the initial residual graph. This gives us some flexibility in using the algorithm, as any capacity matrix representation can be converted into a 
  function easily, and without loosing efficiency for read-access\footnote{Due to technical limitations of our tools, this function 
  cannot depend on heap content, but it can use the efficient functional and pseudo-functional data structures provided by the Isabelle Collection Framework.}. 
  Similarly, our implementation expects an adjacency map as additional parameter, which is then tabulated into an array. This is convenient in our context, where a preprocessing step computes the adjacency map anyway.

@@

  Note that there is still some optimization potential left in the choice of our data structures: 
  For example, the BFS algorithm computes a predecessor map $P$. It then iterates over $P$ to extract the shortest path as a list of edges.
  A subsequent iteration over this list computes the residual capacity, and a final iteration performs the augmentation. 
  This calls for a deforestation optimization to get rid of the intermediate list, and iterate only two times over the predecessor map directly.
%   A first time to compute the bottleneck, and a second time to do the augmentation. 
%   Fortunately, iteration over the shortest path seems not to significantly contribute to the runtime of our implementation, such that 
  We leave this optimization for future work.

@@

  Combining the implementation of the Edmonds-Karp algorithm with the network checker yields our final implementation,
  for which we can export code, and have proved the following theorem: 
  \Snippet{edmonds_karp_correct}
  Note that this theorem is stated as a Hoare triple, using separation logic assertions. There are no preconditions on the input.
  If the algorithm returns \isai{None}, then the edge list was malformed or described a graph that does not satisfy the network assumptions.
  Here, \isai{ln_invar} describes well-formed edge lists, \ie edge lists that have no duplicate edges and only edges with positive capacity,
  and \isai{ln_\<alpha>} describes the mapping from (well-formed) edge lists to capacity matrices.
  If the algorithm returns some number $N$ and residual graph $cf$, then the input was a well-formed edge list that describes a valid network with at most $N$ nodes.
  Moreover, the returned residual graph describes a flow $f$ in the network, which is maximal. 
  As there are no other possible return values, this theorem completely describes the correctness of the algorithm. 
  Note that Isabelle/HOL does not have a notion of execution, thus total correctness of the generated code cannot be expressed. However, the program is phrased in a heap-exception monad, thus introducing some (coarse grained) notion of
  computation. On this level, termination can be ensured, and, indeed, the above theorem implies that all the recursions stated by recursion 
  combinators in the monad must terminate. However, it does not guarantee that we have not injected spurious code equations like $f~x = f~x$, 
  which is provable by reflexivity, but causes the generated program to diverge.

\section{Benchmarking}\label{sec:benchmark}
  We have compared the running time of our algorithm against an unverified reference implementation in Java, taken from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}.
  We have done the comparison on randomly generated sparse and dense networks, the sparse networks having a \emph{density} (${} = \frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$. Note that the maximum density for networks that satisfy our assumptions is $0.5$, as we allow no parallel edges.
  For sparse networks, we varied the number of nodes between $1000$ and $5000$, for dense networks between $1000$ and $1400$.
  The results are shown in Figure~\ref{fig:benchmark}, in a double-logarithmic scale. 
%   \begin{wrapfigure}{o}{.4\textwidth}
%   \vspace{-20pt}
%   \begin{center}
%   \begin{tabular}{|c|c|c|}
%   \hline
%   Name        & SML       & Java   \\
%   \hline\hline
%   1000-sparse & 204       & 15     \\
%   2000-sparse & 1430      & 231    \\
%   3000-sparse & 4928      & 828    \\
%   4000-sparse & 9299      & 2145   \\
%   5000-sparse & 23625     & 4868   \\
%   1000-dense  & 6575      & 4011   \\
%   1100-dense  & 8067      & 5487   \\
%   1200-dense  & 10892     & 7011   \\
%   1300-dense  & 14070     & 9815   \\
%   1400-dense  & 17317     & 12156  \\
%   \hline
%   \end{tabular}
%   \end{center}
%   \vspace{-15pt}
%   \caption{Dijkstra benchmark}\label{tab:benchmark}
%   \vspace{-20pt}
%   \end{wrapfigure}

  \begin{figure}[!h]
  \centering    
  \begin{tikzpicture}[thick,scale=1, every node/.style={scale=1}] %change the scales if you like to reduce the size
    \begin{axis}[
      %title={Benchmark of the Edmonds-Karp algorithm},
        axis x line*= bottom,
        axis y line*= left,
        xmode = log,
        ymode = log,
        xlabel = Number of vertices in the network,
        ylabel = Execution time in seconds, %in milliseconds, modify the scale for lables
        xtick = data,
        ytick = data,
        xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$}, %corresponds to the first column in {SML, Sparse}
        yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
        smooth,
        cycle list name = black white,
        legend style = { 
          at={(0.59,0.4)}, 
            anchor=north west,
            draw=black, 
            fill= white,
            align=left
        }
    ]

    %start data for plotting
      \addplot table {
        1000 204
        2000 1430
        3000 4928
        4000 9299
        5000 23625
      }; \addlegendentry{SML, sparse};

      \addplot table {
        1000 15
        2000 231
        3000 828
        4000 2145
        5000 4868
      }; \addlegendentry{Java, sparse};

      \addplot table {
        1000 6575
        1100 8067
        1200 10892
        1300 14070
        1400 17317
      }; \addlegendentry{SML, dense};

      \addplot table {
        1000 4011
        1100 5487
        1200 7011
        1300 9815
        1400 12156
      }; \addlegendentry{Java, dense};
    %end data for plotting

    \end{axis}
  \end{tikzpicture}
  \caption{Edmonds-Karp benchmark}\label{fig:benchmark}
  \end{figure}

  We observe that, for sparse graphs, the Java implementation is roughly faster by a factor of 6, while for dense graphs, our implementation comes quite close 
  to the Java implementation. We have no full explanation for this phenomenon. Note that the Java implementation operates on flows, while our implementation 
  operates on residual graphs (\cf Section~\ref{sec:impl_res_graph}). Moreover, the Java implementation does not store the augmenting 
  path in an intermediate list, but uses the predecessor map computed by the BFS directly (\cf Section~\ref{sec:impl_data_structures}).

  Finally note that a carefully optimized C++ implementation of the algorithm is only slightly faster than the Java implementation for sparse graphs,
  but roughly one order of magnitude faster for dense graphs. We leave it to future work to investigate this issue, and conclude that we were able to produce
  a reasonably fast verified implementation, without doing any low-level optimizations, but using the code as generated by the Isabelle/HOL code generator.

>>>>

\author{Peter Lammich, S.~Reza Sefidgar}

\institute{Technische Universit\"at M\"unchen, \email{\{lammich,sefidgar\}@in.tum.de}}

@@

\section{Introduction}
Computing the maximum flow of a network is an important problem in graph theory.
Many other problems, like maximum-bipartite-matching, edge-disjoint-paths, circulation-demand, as well as various scheduling and resource allocating problems can be reduced to it.
The Ford-Fulkerson method~\cite{FF56} describes a class of algorithms to solve the maximum flow problem, an important instance being the Edmonds-Karp algorithm~\cite{EK72},
which was one of the first algorithms to solve the maximum flow problem in polynomial time for the general case of real valued capacities.

@@

While there exists another formalization of the Ford-Fulkerson method in Mizar~\cite{Lee05}\footnote{Section~\ref{sec:related_work} provides a detailed discussion}, we are, to the best of our knowledge, the first that verify a polynomial maximum flow algorithm, prove the polynomial complexity bound, or provide a verified executable implementation. Moreover, this paper is a case study on elegantly formalizing algorithms.

@@

To simplify reasoning about the maximum flow problem, we assume that our network satisfies some additional constraints: 1) The source only has outgoing edges while the sink only has incoming edges; 2) If the flow network contains an edge $(u, v)$ then there is no \emph{parallel edge} $(v, u)$ in the reverse direction\footnote{With $u=v$, this also implies that there are no self loops.}; and 3) every vertex of the flow network must be on a path from $s$ to $t$. Note that any network can be transformed to a network with the aforementioned properties and the same maximum flow~\cite{CLRS09}.

@@

The Ford-Fulkerson method is a corollary of this theorem. It is based on a greedy approach: Starting from a zero flow, the value of the flow is iteratively increased until a maximum flow is reached. In order to increase the overall flow value, it may be necessary to redirect some flow, \ie to decrease the flow passed through specific edges. For this purpose the Ford-Fulkerson method defines the residual graph, which has edges both in parallel and reverse to the edges of the network. 
Each edge is labeled by the amount of flow that can be effectively moved along this edge, by either increasing or decreasing the flow on a network edge.
Formally, the residual graph $G_f$ of a flow $f$ is the graph induced by the 
edges with positive labels according to the following labeling function $c_f$:
\[ c_f (u, v) = 
  \begin{cases}
  c (u, v) - f(u, v) \hfill & \text{ if $(u, v) \in E$} \\
  f (v, u) \hfill & \text{ if $(v, u) \in E$} \\  
  0 \hfill & \text{ otherwise} \\
  \end{cases} 
\]

@@

The Ford-Fulkerson method does not specify how to find an augmenting path in the residual graph, and there are several possible implementations with different execution times. The general method is only guaranteed to terminate for networks with rational capacities, while it may infinitely converge against non-maximal flows in the case of irrational edge capacities~\cite{FF56,Zwick95}. When always choosing a \emph{shortest} augmenting path, the number of iterations is bound by $O(VE)$, even for the general case of real valued capacities. Note that we write $V$ and $E$ instead of $|V|$ and $|E|$ for the number of nodes and edges if the intended meaning is clear from the context.
A shortest path can be found by breadth first search (BFS) in time $O(E)$, yielding the Edmonds-Karp algorithm~\cite{EK72} with an overall running time of $O(VE^2)$.

\section{Formalizing the Ford-Fulkerson Method}\label{sec:abs-formalization}
%  GOAL: Present highlights of our formalization, persuade the reader that we have done something substantial.
%    Reader should think: Yeah, they did cool stuff!
%
%  [???* Thematize definition of flows: On Graphs vs. on Networks. You find both in the literature. 
%      We decides for [...]. This is better suited, as is gives nice and elegant proofs, even formally, bla bla bla]
%    
%  Proof of MinCut-MaxFlow: Follows the textbook proof.
%    Uses Isar to even look like a textbook proof, being comprehensible even without running
%      Isabelle/HOL.
%      
%    For example, \isai{augment_flow_presv_cap}.
%      Present formal proof text. Perhaps oppose it to textbook proof?.
%    
%  Abstract Algo looks like pseudocode presented in texbooks.
%    Oppose our algo to textbook pseudocode.
%

@@

In this section, we provide a brief overview of our formalization of the Ford-Fulkerson method. 
In order to develop theory in the context of a fixed graph or network, we use Isabelle's concept of \emph{locales}~\cite{Ballarin:2006:MKM}, which allows
us to define named contexts that fix some parameters and assumptions.
For example, the graph theory is developed in the locale \isai{Graph}, which 
fixes the edge labeling function \isai{c}, and defines the set of edges and nodes based on $c$:
\begin{lstlisting}
  locale Graph = fixes c :: "edge => 'capacity" begin
    definition "E == {(u, v). c (u, v) ~= 0}"
    definition "V == {u. \<exists>v. (u, v) \<in> E \<or> (v, u) \<in> E}"
    [...]
\end{lstlisting}
Moreover, we define basic concepts like (simple, shortest) paths, and provide lemmas to reason about them.

Networks are based on graphs, and add the source and sink node, as well as the network assumptions:
\begin{lstlisting}
  locale Network = Graph + fixes s t :: node
    assumes no_incoming_s: "\<forall>u. (u, s) \<notin> E"
    [...]
\end{lstlisting}
Most theorems presented in this paper are in the context of the \isai{Network} locale.

\subsection{Presentation of Proofs}
% (DONE: not sure about repeating citation of Cormen: I would definitely repeat the citation!) 
Informal proofs focus on the relevant thoughts by leaving out technical details and obvious steps. 
In contrast, a formal proof has to precisely specify each step as the application of some inference rules. 
Although modern proof assistants provide high-level tactics to summarize some of these steps, formal proofs tend to be significantly more 
verbose than informal proofs. Moreover, formal proofs are conducted in the tactic language of the proof assistant, which is often some dialect of ML. 
Thus, many formal proofs are essentially programs that instruct the proof assistant how to conduct the proof. They tend to be inaccessible without a deep knowledge of the used proof assistant, in many cases requiring to replay the proof in the proof assistant in order to understand the idea behind it.

@@

\subsection{Presentation of Algorithms}
In textbooks, it is common to present algorithms in pseudocode, which captures the essential ideas, but leaves open implementation details. As a formal equivalent to pseudocode, we use the monadic programming language provided by the Isabelle Refinement Framework~\cite{LaTu12,La12}. For example, we define the Ford-Fulkerson method as follows:
\Snippet{ford_fulkerson_algo}

% DONE: Introduce obtain-syntax, and unfold find-augmenting-path spec! Or even while-exists loop as combinator?
The code looks quite similar to pseudocode that one would expect in a textbook, but actually is a rigorous formal specification of the algorithm, using nondeterminism to leave open the implementation details (\cf Section~\ref{sec:refinement}). Note that we had to use the available combinators of the Isabelle Refinement Framework, which made the code slightly more verbose than we would have liked. We leave it to future work to define a set of combinators and appropriate syntax that allows for more concise presentation of pseudocode.

@@

In Isabelle/HOL, stepwise refinement is supported by the Isabelle Refinement Framework~\cite{LaTu12,La12}. 
It features a refinement calculus for programs phrased in a nondeterminism monad. 
The monad's type is a set of possible results plus an additional value that indicates a failure:
\begin{lstlisting}
  datatype \<alpha> nres = res \<alpha> set | fail
\end{lstlisting}
The operation \isai{return x} of the monad describes the single result \isai{x}, and the 
operation \isai{bind m f} nondeterministically picks a result from \isai{m} and executes \isai{f} on it. 
If either \isai{m = fail}, or \isai{f} may fail for a result in \isai{m}, the bind operation fails.

We define the \emph{refinement ordering} on \isai{\<alpha> nres} by lifting the subset ordering with \isai{fail} being the greatest element.
Intuitively, \isai{m<=m'} means that \isai{m} is a refinement of \isai{m'}, \ie all possible results of \isai{m} are 
also possible results of \isai{m'}. 
Note that the refinement ordering is a complete lattice, and bind is monotonic. Thus, we can define recursion using a fixed point construction~\cite{Kr10}.
Moreover, we can use the standard Isabelle/HOL constructs for if, let and case distinctions, yielding a fully fledged programming 
language, shallowly embedded into Isabelle/HOL's logic. For simpler usability, we define standard loop constructs (while, foreach), 
a syntax for postcondition specifications, and use a Haskell-like do-notation:
\begin{lstlisting}
  spec P == spec x. P x == res {x. P x}
  do {x <- m; f x} == bind m f
  do {m; m'} == bind m (%_. m')
\end{lstlisting}

Correctness of a program \isai{m} with precondition \isai{P} and postcondition \isai{Q} is expressed as \isai{P ==> m <= spec r. Q r} (or, eta-contracted, just \isai{spec Q}), which
means that, if \isai{P} holds, \isai{m} does not fail and all possible results of \isai{m} satisfy \isai{Q}. Note that we provide different recursion constructs
for partial and total correctness: A nonterminating total correct recursion yields \isai{fail}, which satisfies no specification, even if joined 
with results from other possible runs. On the other hand, a nonterminating partial correct recursion yields \isai|res {}|, which refines any specification 
and disappears when joined with other results.

The Isabelle Refinement Framework also supports data refinement, changing the representation of results according to a \emph{refinement relation}, 
which relates concrete with abstract results: Given a relation \isai{R}, \isai{\<Down>R m} is the largest set of concrete results that are related to an 
abstract result in \isai{m} by \isai{R}. If \isai{m=fail}, then also \isai{\<Down>R m = fail}.

In a typical program development, one first comes up with an initial version \isai{m_0} of
the algorithm and its specification \isai{P,Q}, and shows \isai{P ==> m_0 <= spec Q}. Then, one iteratively provides refined versions \isai{m$_i$} of the algorithm,
proving \isai{m$_i$ <= \<Down>R$_i$ m$_{i-1}$}. Using transitivity and composability of data refinement, one 
gets \isai{P ==> m$_i$ <= \<Down>R$_i$...R$_1$ spec Q}, showing the correctness of the refined algorithm.
If no data refinement is performed, \isai{R$_i$} is set to the identity relation, in which case \isai{\<Down>R$_i$} becomes the identity function.

@@

\section{The Edmonds-Karp Algorithm}\label{sec:edka}
  Specializing the Ford-Fulkerson method to the Edmonds-Karp algorithm requires to show that finding a shortest augmenting path is a refinement of finding any augmenting path, which is straightforward using basic graph theory.

@@

  Formalizing the above intuitive argument was more tricky than it seemed on first glance:
  While it is easy to prove that, in a \emph{fixed graph}, an edge and its reverse cannot both lie on shortest paths, generalizing the argument
  to a graph transformation which may add multiple reverse edges and removes at least one original edge requires some generalization of the statement.
  Note that a straightforward induction on the length of the augmenting path or on the number of flipped edges fails, as, after flipping the first 
  edge, the path is no longer augmenting.

@@

  Having defined the measure function and shown that it decreases on augmentation, it is straightforward to refine the partial correct while loop to a total correct one. Moreover, to make explicit the bound on the number of loop iterations, we instrument the loop with a counter, and assert the upper bound on the counter after the loop.

\section{Refinement to Executable Code}\label{sec:executable}
  In the previous section, we have presented our abstract formalization of the Edmonds-Karp algorithm,
  leaving open how to obtain a shortest augmenting path and how to implement the algorithm. 
  In this section, we outline the further refinement steps that were necessary to obtain an efficient implementation.

@@

  \subsection{Implementing Augmentation}    
  In our abstract formalization, which matches the presentation in Section~\ref{sec:background}, we have formulated augmentation by first defining 
  the residual capacity $c_p$ of the augmenting path. Using $c_p$, we have defined the residual flow $f_p$, which was finally added to the current flow.
  In the refinement to operate on residual graphs, we have refined this to augment the residual graph.
  For the implementation, we compute the residual capacity in a first iteration over the augmenting path, and modify the residual graph 
  in a second iteration. Proving this implementation correct is straightforward by induction on the augmenting path.

@@

  We implement capacities as (arbitrary precision) integer numbers\footnote{Up to this point, the formalization models capacities as \emph{linearly ordered integral domains}, which subsume reals, rationals, and integers.}. Note that an implementation as fixed precision numbers would also be possible,
  but requires additional checks on the network to ensure that no overflows can happen. 
  % TODO: Can we trick the Java reference implementation on integer overflow?

@@

  The input network of the algorithm is represented as a function from network edges to capacities, which is tabulated into an array
  to obtain the initial residual graph. This gives us some flexibility in using the algorithm, as any capacity matrix representation can be converted into a 
  function easily, without loosing efficiency for read-access. 
  Similarly, our implementation expects an adjacency map as additional parameter, which is then tabulated into an array. This is convenient in our context, where a preprocessing step computes the adjacency map anyway.

@@

  Note that there is still some optimization potential left in the choice of our data structures: 
  For example, the BFS algorithm computes a predecessor map $P$. It then iterates over $P$ to extract the shortest path as a list of edges.
  A subsequent iteration over this list computes the residual capacity, and a final iteration performs the augmentation. 
  This calls for a deforestation optimization to get rid of the intermediate list, and iterate only two times over the predecessor map directly.
%   A first time to compute the bottleneck, and a second time to do the augmentation. 
  Fortunately, iteration over the shortest path seems not to significantly contribute to the runtime of our implementation, such that 
  we leave this optimization for future work.

@@

  Combining the implementation of the Edmonds-Karp algorithm with the network checker yields our final implementation,
  for which we can export code, and have proved the following theorem: 
  \Snippet{edmonds_karp_correct}
  Note that this theorem is stated as a Hoare triple, using separation logic assertions. There are no preconditions on the input.
  If the algorithm returns \isai{None}, then the edge list was malformed or described a graph that does not satisfy the network assumptions.
  Here, \isai{ln_invar} describes well-formed edge lists, \ie edge lists that have no duplicate edges and only edges with positive capacity,
  and \isai{ln_\<alpha>} describes the mapping from (well-formed) edge lists to capacity matrices (note that we set \isai{c==ln_\<alpha> el}).
  If the algorithm returns some number \isai{N} and residual graph \isai{cf}, then the input was a well-formed edge list that describes a valid network with at most \isai{N} nodes.
  Moreover, the returned residual graph describes a flow \isai{f} in the network, which is maximal. 
  As the case distinction is exhaustive, this theorem states the correctness of the algorithm. 
  Note that Isabelle/HOL does not have a notion of execution, thus total correctness of the generated code cannot be expressed. However, the program is phrased in a heap-exception monad, thus introducing some (coarse grained) notion of
  computation. On this level, termination can be ensured, and, indeed, the above theorem implies that all the recursions stated by recursion 
  combinators in the monad must terminate. However, it does not guarantee that we have not injected spurious code equations like $f~x = f~x$, 
  which is provable by reflexivity, but causes the generated program to diverge.

\section{Benchmarking}\label{sec:benchmark}
  We have compared the running time of our algorithm against an unverified reference implementation in Java, taken from Sedgewick and Wayne's book on algorithms~\cite{SeWa11}. We have used MLton 20100608~\cite{MLton} and OpenJDK Java 1.7.0\_95, running on a standard laptop machine with a 2.8GHz i7 quadcore processor and 16GiB of RAM.

  We have done the comparison on randomly generated sparse and dense networks, the sparse networks having a \emph{density} (${} = \frac{E}{V(V - 1)}$) of $0.02$, and the dense networks having a density of $0.25$. Note that the maximum density for networks that satisfy our assumptions is $0.5$, as we allow no parallel edges.
  For sparse networks, we varied the number of nodes between $1000$ and $5500$, for dense networks between $1000$ and $1450$.
  The results are shown in Figure~\ref{fig:benchmark}, in a double-logarithmic scale.

  \begin{figure}[!h]
  \begin{tikzpicture}[thick,scale=.7, every node/.style={scale=1}] %change the scales if you like to reduce the size
    \begin{axis}[
      %title={Benchmark of the Edmonds-Karp algorithm},
        axis x line*= bottom,
        axis y line*= left,
        xmode = log,
        ymode = log,
        xlabel = Number of vertices in the network,
        ylabel = Execution time in milliseconds, %in milliseconds, modify the scale for lables
        xtick = {0,1000,...,5000},
%         ytick = data,
        xticklabels = {$1000$, $2000$, $3000$, $4000$, $5000$}, %corresponds to the first column in {SML, Sparse}
%         yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
        smooth,
        cycle list name = black white,
        legend style = { 
          at={(0.59,0.4)}, 
            anchor=north west,
            draw=black, 
            fill= white,
            align=left
        }
    ]

    %start data for plotting
      \addplot table {
        1000 27   
        1500 124  
        2000 299  
        2500 665  
        3000 1335 
        3500 2526 
        4000 3111 
        4500 6917 
        5000 7743 
        5500 10452    
      }; \addlegendentry{SML, sparse};

      \addplot table {
        1000 16   
        1500 49   
        2000 168  
        2500 453  
        3000 842  
        3500 1524 
        4000 2136 
        4500 3438 
        5000 4734 
        5500 6369     
      }; \addlegendentry{Java, sparse};
    %end data for plotting
    \end{axis}
  \end{tikzpicture}
%
  \begin{tikzpicture}[thick,scale=.7, every node/.style={scale=1}] %change the scales if you like to reduce the size
    \begin{axis}[
      %title={Benchmark of the Edmonds-Karp algorithm},
        axis x line*= bottom,
        axis y line*= left,
        xmode = log,
        ymode = log,
        xlabel = Number of vertices in the network,
        ylabel = Execution time in milliseconds, %in milliseconds, modify the scale for lables
        xtick = {0,1000,1100,...,1400},
        ytick = {0,100,1000,10000},
        xticklabels = {$1000$, $1100$, $1200$, $1300$, $1400$}, %corresponds to the first column in {SML, Sparse}
%         yticklabels = {$0.2$, $1.4$, $4.9$, $9.2$, $23.6$}, %corresponds to the second column in {SML, Sparse}
        smooth,
        cycle list name = black white,
        legend style = { 
          at={(0.59,0.4)}, 
            anchor=north west,
            draw=black, 
            fill= white,
            align=left
        }
    ]

    %start data for plotting
      \addplot table {
        1000 3328 
        1050 3345 
        1100 4841 
        1150 5036 
        1200 5840 
        1250 6858 
        1300 7651 
        1350 8950 
        1400 8398 
        1450 10769     
      }; \addlegendentry{SML, dense};

      \addplot table {
        1000 4141 
        1050 4778 
        1100 5465 
        1150 6248 
        1200 7128 
        1250 8137 
        1300 9864 
        1350 11008
        1400 11902
        1450 13365     
      }; \addlegendentry{Java, dense};
    %end data for plotting
    \end{axis}
  \end{tikzpicture}
